Adaptive Shrinkage and False Discovery Rates by Laplace Approximation

### 2014-07-13 talks/ASC2014.Rmd
add ASC talk

% False Discovery Rates, A New Deal
% Matthew Stephens
% 2014/2/24

@@

```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(cache=TRUE,autodep=TRUE,warning=FALSE)
dep_auto()
```

# Before we get started: Getting Organized

@@

- Buy a notebook; bring it to meetings; make notes! 
- Come to meetings with a written agenda. 
- While doing research, record what you did and what the outcome was.
- Use version control ([git](http://git-scm.com)) and internet repositories ([bitbucket](http://www.bitbucket.org), [github](http://www.github.com)) to organize notes, code, etc.
- Use *knitr* to help make your research reproducible.

@@

- Apply `qvalue` 
to identify findings ``significant" at a given False Discovery Rate.

@@

@@

```{r, echo=FALSE}
  #Set up Analysis of Hedenfalk BRCA data
  setwd("~/Documents/git/ash/paper/Rcode")

  altcol="cyan" #colors to use
  nullcol="blue" 
  nc=40 #number of bins in histograms
  ncz = 100 # number of bins in z score histograms

  hh = read.table("../../data/nejm_brca_release_edit.csv",sep=",",skip=3)
  subset = apply(hh, 1, max)<20
  hh = hh[subset,]

  labs = read.table("../../data/nejm_brca_release_edit.csv",sep=",",skip=1,nrows=1)
  labs = 1*(labs=="BRCA1") + 2 * (labs=="BRCA2")

  hh.betahat = apply(hh[,labs==1],1,mean) - apply(hh[,labs==2],1,mean)
  n1 = sum(labs==1)
  n2 = sum(labs==2)
  hh.sebetahat = sqrt(apply(hh[,labs==1],1,var)/n1 + apply(hh[,labs==2],1,var)/n2)
  hh.zscore = hh.betahat/hh.sebetahat
  hh.pval = pchisq(hh.zscore^2,df=1,lower.tail=F)

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

  hh.q = qvalue(hh.pval)
  abline(h=hh.q$pi0,col=nullcol,lwd=2)

  hh.hist$density=rep(hh.q$pi0,length(hh.hist$density))  
  #hh.hist$counts=rep(hh.q$pi0*length(hh.pval)/nc,length(hh.hist$counts)) 
  plot(hh.hist,add=TRUE,col=nullcol,freq=FALSE)
```

Data from Hedenfalk et al, comparing BRCA1 vs BRCA2 expression.

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

  hh.q = qvalue(hh.pval)
  abline(h=hh.q$pi0,col=nullcol,lwd=2)

  hh.hist$density=rep(hh.q$pi0,length(hh.hist$density))  
  #hh.hist$counts=rep(hh.q$pi0*length(hh.pval)/nc,length(hh.hist$counts)) 
  plot(hh.hist,add=TRUE,col=nullcol,freq=FALSE)

  abline(v=0.1,lwd=2,col=2)

  text(0.05,1.2,labels="A",col=2,cex=1.2)  
  text(0.05,0.4,labels="B",col=2,cex=1.2)  
  text(0.6,3,labels=paste0("FDR = B/(A+B) =  ",round(hh.q$pi0*0.1*length(hh.pval)/sum(hh.pval<0.1),2)),cex=1.2)
```

# Example: fdr estimation

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

# Example: fdr estimation

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

  require(fdrtool)
  hh.gren = grenander(ecdf(hh.pval))
  abline(h=min(hh.gren$f.knots),col=nullcol,lwd=2)  
  lines(hh.gren$x.knots,hh.gren$f.knots,lwd=2)
```

# Example: fdr estimation

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

  require(fdrtool)
  hh.gren = grenander(ecdf(hh.pval))
  abline(h=min(hh.gren$f.knots),col=nullcol,lwd=2)  
  lines(hh.gren$x.knots,hh.gren$f.knots,lwd=2)
  abline(v=0.1,lwd=2,col=2)
  text(0.1,0.9,labels="a",col=2,cex=1)  
  text(0.1,0.34,labels="b",col=2,cex=1.2)  
  text(0.6,3,labels=paste0("fdr = b/(a+b) =  ",round(min(hh.gren$f.knots)/approx(hh.gren$x.knots,hh.gren$f.knots,0.1)$y,2)),cex=1.2)
```

# FDR problem 1: different measurement precision

- If some effects are measured very imprecisely, those tests ``lack power"
and simply add noise

- In particular, such tests increase the estimated number of nulls, and increase
the FDR for other tests

- It would seem preferable to simply ignore the tests with very low precision. Summarizing each test by a $p$ value (or $Z$ score) loses the information about precision.

@@

```{r, echo=FALSE}
setwd("~/Documents/git/ash/talks/")
## load Poisson_binomial and ash functions 
source("../../stat45800/code/Rcode/PoissonBinomial.funcs.R")  
#source("../Rcode/ash.R")

x = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(x[,2:5])
x = x[xx>0,]
xx = xx[xx>0]
```

@@

```{r, echo=FALSE}

cc = x[, 2:5]

g = c(-1, -1, 1, 1)
ngene = dim(x)[1]

cc.assoc = counts.associate(cc, g, 1)
zdat.ash = cc.assoc$zdat.ash
zdat = cc.assoc$zdat

# two-sided test
ttest.pval = function(t, df) {
    pval = pt(t, df = df, lower.tail = T)
    pval = 2 * ifelse(pval < 0.5, pval, 1 - pval)
    return(pval)
}

@@

highxp = xx>1000 # select high expressed genes
pval.high = pval[highxp]
qval.high = qvalue(pval.high)
cc.assoc.high = counts.associate(cc[highxp,],g,1)
zdat.ash.high = cc.assoc.high$zdat.ash
zdat.high = cc.assoc.high$zdat

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
h=hist(pval,prob=TRUE,ylim=c(0,4),main="p values, all genes",xlab="p value",breaks=seq(0,1,length=21),col=altcol)
abline(h=qval$pi0,col=nullcol,lwd=2)
h$density=rep(qval$pi0,length(h$density))  
plot(h,col=nullcol,add=TRUE,freq=FALSE)
abline(v=0.05,lwd=3)
text(0.6,3,labels=paste0("FDR = ",round(max(qval$qval[pval<0.05]),2)),cex=1.2)
```

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
h=hist(pval[xx<1000],prob=TRUE,ylim=c(0,4),main="p values, low count genes",xlab="p value",breaks=seq(0,1,length=21),col=altcol)
qval.low=qvalue(pval[xx<1000])
pval.low=pval[xx<1000]
abline(h=qval.low$pi0,col=nullcol,lwd=2)
h$density=rep(qval.low$pi0,length(h$density))  
plot(h,col=nullcol,add=TRUE,freq=FALSE)
abline(v=0.05,lwd=3)
#text(0.6,3,labels=paste0("FDR = ",round(max(qval.low$qval[pval.low<0.05]),2)),cex=1.2)
```

# Higher count genes, more power
```{r, echo=FALSE, fig.height=4,fig.cap=""}

h=hist(pval.high,prob=TRUE,ylim=c(0,4),main="p values, high count genes",xlab="p value",col=altcol)

abline(h=qval.high$pi0,col=nullcol,lwd=2)
h$density=rep(qval.high$pi0,length(h$density))  
plot(h,col=nullcol,add=TRUE,freq=FALSE)
abline(v=0.05,lwd=3)
#text(0.6,3,labels=paste0("FDR = ",round(max(qval.high$qval[pval.high<0.05]),2)),cex=1.2)
#qval.high$q[pval.high<0.05] gives FDR for p<0.05
```

# FDR problem 1: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[xx>1000],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# FDR problem 1: Summary

- Analyzing $p$ values or $Z$ scores doesn't fully account 
for measurement precision.

# Problem 2: The Zero Assumption (ZA)

- The ZA allows us to estimate the null proportion, $\pi_0$, using the density of $p$ values near 1 (or $Z$ scores near 0).

# Problem 2: The ZA

- The ZA seems initially natural.

- However, it turns out to imply unrealistic assumptions about the distribution of non-zero effects.

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  nc=40 #number of bins in histogram
  hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main="Distribution of p values",nclass=nc,col=altcol)

  hh.q = qvalue(hh.pval)
  abline(h=hh.q$pi0,col=nullcol,lwd=2)

  hh.hist$density=rep(hh.q$pi0,length(hh.hist$density))  
  #hh.hist$counts=rep(hh.q$pi0*length(hh.pval)/nc,length(hh.hist$counts)) 
  plot(hh.hist,add=TRUE,col=nullcol,freq=FALSE)

  abline(v=0.1,lwd=2,col=2)

  text(0.05,1.2,labels="A",col=2,cex=1.2)  
  text(0.05,0.4,labels="B",col=2,cex=1.2)  
  text(0.6,3,labels=paste0("FDR = B/(A+B) =  ",round(hh.q$pi0*0.1*length(hh.pval)/sum(hh.pval<0.1),2)),cex=1.2)
```

# Implied distribution of Z scores under alternative (fdrtool)

```{r, echo=FALSE,fig.height=4,fig.cap=""}
#plot a histogram of z scores, highlighting the alternative distribution
#of z scores that is implied by localfdr values lfdr.
  nullalthist = function(z,lfdr,...){
    h=hist(z, freq=FALSE,col=nullcol,nclass=ncz,...)
    avlfdr = unlist(lapply(split(lfdr,cut(z,h$breaks),drop=FALSE),mean))
    h$density = (1-avlfdr) * h$density
    plot(h,add=TRUE,col=altcol,freq=FALSE)
  }

  nullalthist(hh.zscore,hh.fdrtool$lfdr,main="fdrtool")  
```

# Implied distribution of Z scores under alternative (locfdr)

```{r, echo=FALSE,fig.height=4,fig.cap=""}

  nullalthist(hh.zscore,hh.locfdr$fdr,main="locfdr")
```

# Implied distribution of Z scores under alternative (mixfdr)

```{r, echo=FALSE,fig.height=4,fig.cap=""}

  nullalthist(hh.zscore,hh.mixfdr$fdr,main="mixfdr")
  dev.off()
```

@@

- By summarizing each observation by a $Z$ score or $p$ value, 
standard fdr tools ignore precision of different measurements

- Standard tools make the ZA, which implies actual effects have a (probably unrealistic) bimodal distribution. [and tends to overestimate $\pi_0$, losing power]

- Also standard tools focus only on zero vs non-zero effects. (eg what if we would
like to identify genes that have at least a 2-fold change?)

@@

- Eg Efron assumes that the $Z$ scores come from a mixture of null, and alternative:
$$Z_j \sim f_Z(.) = \pi_0 N(.;0,1) + (1-\pi_0) f_1(.)$$
where $f_1$ is to be estimated from the data.

@@

- $\text{fdr}(Z) \approx \pi_0 N(Z; 0,1)/ f_Z(Z)$

# FDR: The New Deal

@@

- *Incorporate precision* of each observation $\hat\beta$ into the likelihood.
Specifically, approximate likelihood for $\beta_j$ by a normal: 
$$L(\beta_j) \propto \exp(-0.5 (\beta_j - \hat\beta_j)^2/s_j^2).$$
[From $\hat\beta_j \sim N(\beta_j, s_j)$]

- fdr given by $$p(\beta_j =0 | \hat\beta_j) = \pi_0 p(\hat\beta_j | \beta_j=0)/p(\hat\beta_j)$$

@@

- A convenient way to model $g$ is by a mixture of 0-centered
normal distributions: 
$$g(\beta; \pi) = \sum_{k=1}^K \pi_k N(\beta; 0, \sigma^2_k)$$

- Estimating $g$ comes down to estimating $\pi$. Joint estimation of $\pi_0,\pi$ easy by maximum likelihood (EM algorithm) or variational Bayes.

- By allowing $K$ large, and $\sigma_k$ to span a dense grid of values,
we get a fairly flexible unimodal symmetric distribution.

@@

- If allow a very large number of uniforms this provides the non-parametric mle for $g$; cf Grenander 1953; Campy + Thomas.

@@

# Issue: identifiability of $\pi_0$

- For estimating False Discoveries, we are asking whether $\beta_j = 0$.

- However, the data cannot distinguish between $\beta_j = 0$ and $\beta_j$ "very
small"

- As a result $\pi_0$ is formally unidentifiable. Eg data can never rule out $\pi_0=0$.

# Issue: identifiability of $\pi_0$

- The Zero assumption (ZA) solves the identifiability problem by assuming that
there *are* no $\beta_j$ near zero!

- The ZA makes $\pi_0$ identifiable.

- Another view is that the estimate of $\pi_0$ under ZA will systematically tend to overestimate $\pi_0$, and so is ``conservative".

- That is it provides an ``upper bound" on $\pi_0$

# Identifiability of $\pi_0$: Solution 1

- We replaced the ZA with the unimodal assumption on g.

- This does not make $\pi_0$ identifiable, but it does effectively provide an upper bound on $\pi_0$.

- Indeed, we saw that when we estimated $\pi_0$ under the ZA the data
then contradicted the unimodal assumption on g. Thus the upper bound is
more conservative than under ZA.

- In practice, implement upper bound by using penalized likelihood that encourages $\pi_0$ to be as big as possible.

# Illustration: Simulated Example

```{r,echo=FALSE,include=FALSE}
load("../paper/Rcode/sim1_out.RData") #load simulation results
source("../paper/Rcode/plot_pi0.R")
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
plot_pi0(list(simres1))
```

# Example: BRCA data

```{r, echo=FALSE,include=FALSE}
hh.ash = ash(hh.betahat,hh.sebetahat)
hh.ashz = ash(hh.zscore,1, pointmass=TRUE, prior="nullbiased")
hh.ash.hu = ash(hh.betahat, hh.sebetahat,mixcompdist="halfuniform")
```

```{r, echo=FALSE, include=FALSE}
  hh.ash.fdr = ash(hh.zscore,1,pointmass = TRUE,prior="nullbiased")
  hh.ash.shrink = ash(hh.zscore,1)
```

# Recall Problem: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.mixfdr$fdr,main="mixfdr")
```

# Problem Fixed: distribution of alternative Z values unimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.ashz$ZeroProb,main="ash")
```

# BRCA1: Compare $\pi_0$ estimates
```{r}
  round(c(hh.fdrtool$param[3],
  hh.locfdr$fp0[1,3],
  hh.mixfdr$pi[1],
  hh.ashz$fitted.g$pi[1]),2)
```

# BRCA1: Compare number significant at fdr<0.05

```{r}
  c(sum(hh.fdrtool$lfdr<0.05),
  sum(hh.locfdr$fdr<0.05),
  sum(hh.mixfdr$fdr<0.05),
  sum(hh.ashz$ZeroProb<0.05))
```

# Identifiability of $\pi_0$ and the False Sign Rate

- Identifiability of $\pi_0$ is primarily an issue if we insist on asking question is $\beta_j=0$?

- How about we change focus: assume *none* of the $\beta_j$ are zero ("one group approach"), and ask for which $\beta_j$ are we confident about the sign (Gelman et al, 2012).

- Positive and negative effects are often treated differently in practice anyway.

- That is we replace fdr with False Sign Rate (fsr), the probability that if we say an effect is positive (negative), it is not.

- Example: suppose we estimate that $\Pr(\beta_j<0)=0.975$ and $\Pr(\beta_j>0)=0.025$. Then we report $\beta_j$ as a ``(negative) discovery", and estimate its fsr as 0.025.

# The fsr is more robust than fdr

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
source("../paper/Rcode/plot_lfsr.R")
  plot_lfsr(list(simres1),0.1,ptype="lfdr")
```

# The fsr is more robust than fdr

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  plot_lfsr(list(simres1),0.1,ptype="lfsr")
```

# Estimation and Shrinkage

- Besides allowing one to estimate fdr and fsr, 
this approach also provides a full posterior distribution for each $\beta_j$.

- So for example we can easily compute fdrs for discoveries other than ``non-zero" (eg compute $\Pr(\beta_j > 2 | \hat\beta_j)$).

- And use it to obtain point estimates and credible intervals for each $\beta_j$, taking account of information from all the other $\beta_j$.

- Because $f(\beta)$ is unimodal, the point estimates will tend to be ``shrunk" towards the overall mean (0).

- Because $f(\beta)$ is estimated from the data, the amount
of shrinkage is adaptive to the data. And because of the role of $s_j$, the amount of shrinkage adapts to the information on each gene.

@@

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat[3,],main="Raw effect size estimates",xlab="betahat",xlim=c(-2,2),prob=T,ylim=c(0,3))
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat[3,],main="Raw effect size estimates",xlab="betahat",xlim=c(-2,2),prob=T,ylim=c(0,3))
t=seq(-2,2,length=201)
lines(t,dnorm(t,0,sd=0.032), col=2, lwd=2)
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat.ash$PosteriorMean,main="Shrunken estimates",xlab="shrunk betahat",xlim=c(-0.1,0.1),prob=T)
lines(t,dnorm(t,0,sd=0.032), col=2, lwd=2)
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
plot(zdat[3,],zdat.ash$PosteriorMean,main="Estimates vs Shrunken estimates",xlab="betahat",ylab="shrunk betahat",xlim=c(-2,2),ylim=c(-0.1,.1),prob=T)
abline(a=0,b=1,col=2)
```

@@

- But by using two numbers ($\hat\beta,s$) instead of one ($p$ values or $z$ scores) precision of different measurementscan be better accounted for.

@@

- False Sign Rate is more robust to assumptions, and perhaps
therefore preferable, than False Discovery Rate.

@@

- Currently applying it to wavelet shrinkage applications.

# Guarantees?

- ``I think you have some nice ideas. How will you convince
people to use them?" (C Morris)

- Theory anyone?

@@

- Incorporate $t$ likelihood as well as normal.

- Incorporate shrinkage of variances and not just means.

- Extend to allow $g(\cdot;\pi)$ to depend on covariates $X$.

- Extend to allow for correlations in the measured $\hat\beta_j$.

@@

- to the several postdoctoral researchers and students
who have worked with me on related topics.
- Including Scott Powers, Mengyin Lu, Tian Sen, Wei Wang, Zhengrong Xing.

# Reproducible research

- This document is produced with **knitr**, **Rstudio** and **Pandoc**.

- For more details see my `stephens999/ash` repository at `http://www.github.com/stephens999/ash`

- Website: `http://stephenslab.uchicago.edu`

@@

>>>>

% False Discovery Rates, A New Deal
% Matthew Stephens
% 2014/7/8

@@

```{r setup, include=FALSE}
  #set global chunk options
  opts_chunk$set(cache=TRUE,autodep=TRUE,warning=FALSE)
  dep_auto()
```

# Before we get started: Let's get Organized!

@@

- Buy a notebook; bring it to meetings; make notes! 
- Come to meetings with a written agenda. 
- While doing research, record what you did and what the outcome was.
- Use version control ([git](http://git-scm.com)) and internet repositories ([bitbucket](http://www.bitbucket.org), [github](http://www.github.com)) to organize notes, code, etc.
- Use *knitr* to help make your research reproducible.
- Talk about the tools you find useful!

@@

- Apply `qvalue` 
to identify findings ``significant" at a given FDR.

@@

```{r, echo=FALSE}
  #simple simulated example
  ncz = 100 # number of bins in z score histograms
  set.seed(111)
  hh.betahat = rnorm(1000,0,2)
  hh.sebetahat = 1
  hh.zscore = hh.betahat/hh.sebetahat
  hh.pval = pchisq(hh.zscore^2,df=1,lower.tail=F)
  hh.ash = ash(hh.betahat,hh.sebetahat, method="fdr")
  hh.q = qvalue(hh.pval)
```

```{r, echo=FALSE}
  plot_FDReg_hist = function(hh.pval,pi0,nc=40,nullcol="blue",altcol="cyan",type=4,title="Distribution of p values",...){
    hh.hist=hist(hh.pval,freq=FALSE,xlab="p value",main=title,nclass=nc,col=altcol,...)
    if(type>1){
      abline(h=pi0,col=nullcol,lwd=2)

      hh.hist$density=rep(pi0,length(hh.hist$density))  
      #hh.hist$counts=rep(hh.q$pi0*length(hh.pval)/nc,length(hh.hist$counts)) 
      plot(hh.hist,add=TRUE,col=nullcol,freq=FALSE)
    }
    if(type>2){
    abline(v=0.1,lwd=2,col=2)
    }
    if(type>3){
      text(0.05,1.2,labels="A",col=2,cex=1.2)  
      text(0.05,0.4,labels="B",col=2,cex=1.2)  
      text(0.6,3,labels=paste0("FDR = B/(A+B) =  ",round(pi0*0.1*length(hh.pval)/sum(hh.pval<0.1),2)),cex=1.2)
    }
  }
```

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=1)                 
```

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=2)                 
```

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=3)                 
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=4)                 
```

# Is this an important problem?

- The original paper introducing FDR (Benjamini and Hochberg, 1995) has
been cited 21,787 times (May 2014) according to Google Scholar.

- That is three times a day for the last 19 years!

# Problem 1: The Zero Assumption (ZA)

- Seems initially natural.

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=4)                 
```

# Implied distribution of Z scores under alternative

```{r, echo=FALSE,fig.height=4,fig.cap=""}
#plot a histogram of z scores, highlighting the alternative distribution
#of z scores that is implied by localfdr values lfdr.
  nullalthist = function(z,lfdr,nullcol="blue",altcol="cyan",...){
    h=hist(z, freq=FALSE,col=nullcol,nclass=ncz,...)
    avlfdr = unlist(lapply(split(lfdr,cut(z,h$breaks),drop=FALSE),mean))
    h$density = (1-avlfdr) * h$density
    plot(h,add=TRUE,col=altcol,freq=FALSE)
  }

  nullalthist(hh.zscore,hh.fdrtool$lfdr)  
```

# FDR problem 2: different measurement precision

- In some cases the measurement precisions differ among units

- Eg Expression levels of low-expressed genes have less precision than high-expressed genes

- If some effects are measured less precisely than others, those tests ``lack power"
and dilute signal, increasing FDR

@@

```{r, echo=FALSE}
x = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(x[,2:5])
x = x[xx>0,]
xx = xx[xx>0]
```

@@

```{r, echo=FALSE,include=FALSE}

load("../scratchwork/Smemo_qbinom_zdat.RData")
# two-sided test
  ttest.pval = function(t, df) {
    pval = pt(t, df = df, lower.tail = T)
    pval = 2 * ifelse(pval < 0.5, pval, 1 - pval)
    return(pval)
}

@@

highxp = xx>1000 # select high expressed genes
pval.high = pval[highxp]
qval.high = qvalue(pval.high)

zdat.ash = ash(zdat[3,],zdat[4,],df=2,method="fdr",mixcompdist="halfuniform")
zdat.ash.high = ash(zdat[3,highxp],zdat[4,highxp],df=2,method="fdr",mixcompdist="halfuniform")

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
qval= qvalue(pval)
plot_FDReg_hist(pval,pi0=qval$pi0,ylim=c(0,4))
```

@@

```{r, echo=FALSE,fig.height=4,fig.cap=""}
qval.low=qvalue(pval[!highxp])
pval.low=pval[!highxp]
plot_FDReg_hist(pval.low,pi0=qval.low$pi0,ylim=c(0,4))
```

# Higher count genes, more power
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval.high,pi0=qval.high$pi0,ylim=c(0,4))
```

# Low-count genes dilute signal at high-count genes
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval,pi0=qval$pi0,ylim=c(0,4))
```

# FDR problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

@@

Standard tools are unduly conservative.

- The ZA, which implies actual effects have a (probably unrealistic) bimodal distribution; causes overestimate of $\pi_0$, losing power.

- By focussing on $p$ values, low-precision measurements can dilute high-precision measurements.

@@

- Eg Efron assumes that the $Z$ scores come from a mixture of null, and alternative:
$$Z_j \sim f_Z(.) = \pi_0 N(.;0,1) + (1-\pi_0) f_1(.)$$
where $f_1, \pi_0$ are to be estimated from the data.

@@

- Once $f_1$ and $\pi_0$ estimated, FDR calculations are straightforward.

# FDR: A New Deal

@@

- *Incorporate precision* of each observation $\hat\beta$ into the likelihood.
Specifically, approximate likelihood for $\beta_j$ by a normal: 
$$L(\beta_j) \propto \exp(-0.5 (\beta_j - \hat\beta_j)^2/s_j^2).$$
[From $\hat\beta_j \sim N(\beta_j, s_j)$]
Or, better, use a $t$ likelihood if $s_j$ estimated using few observations.

@@

- A convenient way to model $g$: mixture of 0-centered
normal distributions: 
$$g(\beta; \pi) = \sum_{k=1}^K \pi_k N(\beta; 0, \sigma^2_k)$$

- Estimating $g$ comes down to estimating $\pi$. Joint estimation of $\pi_0,\pi$ easy by maximum likelihood (EM algorithm).

- By allowing $K$ large, and $\sigma_k$ to span a dense grid of values,
we get a flexible unimodal symmetric distribution.

@@

- If allow a very large number of uniforms this provides the non-parametric mle for $g$; cf Grenander 1953; Cordy + Thomas 1997.

@@

# Adaptive Shrinkage

- This approach actually provides a full posterior distribution for each $\beta_j$.

- So easy to obtain point estimates and credible intervals.

- Because $g(\beta)$ is unimodal, the point estimates (and CIs) will tend to be ``shrunk" towards the overall mean (0).

- Because $g(\beta)$ is estimated from the data, the amount
of shrinkage is adaptive to signal in the data. And because of the role of $s_j$, the amount of shrinkage adapts to the information on each observation.

@@

# Recall Problem 1: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.fdrtool$lfdr,main="qvalue")  
```

# Problem Fixed: distribution of alternative Z values unimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.ash$lfdr,main="ash")
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,hh.q$pi0,type=4,title="qvalue")                 
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot_FDReg_hist(hh.pval,get_pi0(hh.ash),type=4,title="ash")                 
```

# Recall Problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# Problem Fixed: incorporating precision reduces influence of low-count genes
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(zdat.ash.high$qvalue,zdat.ash$qvalue[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# A new problem: an embarrassment of riches

- If the null is mostly false, the new approach can  provide unsettling results

- The FDR can be small for all observations, even those with $p \approx 1$!

- In the illustrative example, the maximum $q$ value is `r round(max(hh.ash$qvalue),2)`

# Perhaps we didn't really understand the question?

- Problem arises only if we insist on asking question ``is $\beta_j=0$?"

- Given enough signal, we become convinced that very few of the $\beta_j=0$

- But for some $\beta_j$ we still may have little information about actual value

- Suggests a change of focus: ask for which $\beta_j$ are we confident about the sign (cf Gelman et al, 2012).

# The False Sign Rate

- Suggestion: replace FDR with local false sign rate (lfsr), the probability that if we say an effect is positive (negative), it is not.

- Example: suppose we estimate that $\Pr(\beta_j<0)=0.95, \Pr(\beta_j=0) = 0.025$ and $\Pr(\beta_j>0)=0.025$. Then we report $\beta_j$ as a ``(negative) discovery", and estimate its lfsr as 0.05.

# Even with many signals, large $p$ values have high lfsr

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  #res=data.frame(p=hh.pval, lfsr=hh.ash$lfsr, qvalue = hh.ash$qvalue)
  library("reshape2")
  res=data.frame(p=pval, lfsr=zdat.ash$lfsr, qvalue = zdat.ash$qvalue)
  res.melt = melt(res, id.vars=c("p"),variable.name="Measure")
  cbbPalette <- c("#000000", "#D55E00", "#CC79A7")
  labels = c('qvalue','lfsr')
  breaks = c("qvalue","lfsr")

  pp= ggplot(data=res.melt,aes(p,value,color=Measure)) +geom_point(shape=1) +
      geom_abline(colour = "black") +
        xlab("p-value") +
        ylab("lfsr/qvalue")

  print(pp +scale_y_continuous(limits=c(0,1)) +
          scale_x_continuous(limits=c(0,1))  +
           scale_colour_manual(values=cbbPalette,breaks=breaks,labels=labels) +
          coord_equal(ratio=1))

# Advantages of False Sign Rate

- The FDR is typically not actually identifiable from data.

- This is because data cannot distinguish between $\beta_j = 0$ and $\beta_j$ "very
small". So $\pi_0$ is not identifiable, and FDR is very sensitive to $\pi_0$.

- So methods for estimating $\pi_0$ and FDR, including those presented here, are designed to be ``conservative" (i.e. overestimate the FDR).

- The False Sign Rate is much less senstive to $\pi_0$, and hence more identifiable from data!

# Simulated example: $\pi_0$ not identifiable.

```{r,echo=FALSE,include=FALSE}
load("../paper/Rcode/sim1.RData") #load simulation results
source("../paper/Rcode/plot_pi0.R")
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
plot_pi0(list(simres1a))
```

# Simulated Example: so fdr not identifiable

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
source("../paper/Rcode/plot_lfsr.R")
  plot_lfsr(list(simres1a),0.1,ptype="lfdr")
```

# Simulated Example: fsr much more identifiable

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  plot_lfsr(list(simres1a),0.1,ptype="lfsr")
```

@@

- But by using two numbers ($\hat\beta,s$) instead of one ($p$ values or $z$ scores) varying precision of 
measurements is better accounted for.

@@

- False Sign Rate preferable to False Discovery Rate: more identifiable, and
better representation of information in data for ``high-signal" situations.

@@

- E.g. Currently applying it to wavelet shrinkage applications.

@@

- Incorporate shrinkage of variances and not just means. (e.g. ``moderated $t$ test", Smyth et al)

- Allow $g(\cdot;\pi)$ to depend on covariates $X$.

- Allow for correlations in the measured $\hat\beta_j$.

@@

- to the developers of **R**, **knitr**, **Rstudio** and **Pandoc**.

- to the several postdoctoral researchers and students
who have worked with me on related topics.

- Including Ester Pantaleo, Scott Powers, Mengyin Lu, Sen Tian, Wei Wang, Zhengrong Xing.

- NHGRI for funding.

- `ashr` package: `http://www.github.com/stephens999/ash`

@@

# Adaptive Shrinkage of point estimates

- Recall idea: amount of shrinkage depends on measurement precision, $s_j$.

# Adaptive Shrinkage of point estimates

```{r, fig.height=4,echo=FALSE,fig.cap=""}
  res=data.frame(betahat =zdat[3,], Estimate=zdat.ash$PosteriorMean,highxp=highxp)
  cbbPalette <- c("#000000", "#D55E00", "#CC79A7")
  labels = c('betahat','Estimate')
  breaks = c("betahat","Estimate")

  pp= ggplot(data=res,aes(betahat,Estimate,col=as.factor(highxp))) +geom_point(shape=1) +
      geom_abline(colour = "black") +
        xlab("betahat") +
        ylab("shrunk betahat")
